---
title: "Characterising rocky intertidal biodiversity using environmental DNA metabarcoding from local to national scales"
format: html
embed-resources: true
editor: visual
author: Simons D-L, Hipperson H, Webb TJ, Spencer M, Mieszkowska N
chunk_output_type: inline
toc: true
message: false
warning: false
---

## Introduction

This pipeline replicates all analyses in Simons *et al* (2025) "Characterising rocky intertidal biodiversity using environmental DNA metabarcoding from local to national scales".

## Load packages

```{r}
#| label: load-packages
#| echo: true
#| results: false

packages <- c("devtools",
              "qiime2R", #github download
              "microbiome", #github download
              "tidyverse",
              "vegan",
              "ggforce",
              "phyloseq",
              "ANCOMBC",
              "DT",
              "ranacapa", #for ggrare
              "plyr",
              "reshape2",
              "patchwork",
              "microViz",
              "microeco",
              "eulerr",
              "microbiomeutilities",
              "ggtreeExtra", #github
              "mgcv",
              "cowplot",
              "geosphere",
              "MiscMetabar",
              "ggVennDiagram",
              'lme4',
              "car",
              "ggsignif",
              "emmeans",
              "ggeffects",
              "gridExtra",
              "merDeriv", 
              "ggrepel",
              "sf")

new.packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

invisible(lapply(packages, library, character.only = TRUE))
```

## Importing data and tidying

Let's load our phyloseq object and long data, which have already been tidied, decontaminated, filtered for target taxa, and checked for quality control through [this pipeline](https://github.com/dinaleighsimons/intertidal-eDNA-formatting-and-quality-control).

```{r}
#| label: load-data

load("Input_Data/phylo_rocky_eDNA.RData")
phylo_rocky_eDNA_long <- read.csv("Input_data/phylo_rocky_eDNA_long.csv")
```

Let's separate data sets for each primer.

```{r}
#| label: split-data-by-primer

phylo_rocky_eDNA_CO1 <- ps_filter(phylo_rocky_eDNA, primer == "CO1")
phylo_rocky_eDNA_CO1

phylo_rocky_eDNA_18S <- ps_filter(phylo_rocky_eDNA, primer == "18S")
phylo_rocky_eDNA_18S
```

Some of our data analysis (PERMANOVAs) is completed in PRIMER7, so we need to export elements of the `phylo` elements to use outside of R.

```{r}
#| label: extract-data-for-PRIMER

#all
meta_target <- data.frame(sample_data(phylo_rocky_eDNA))
write.csv(meta_target, file = "Processed_Data/PRIMER7_format/meta_target.csv") 

meta_target_t <- t(meta_target)
write.csv(meta_target_t, file = "Processed_Data/PRIMER7_format/meta_target_t.csv") 

taxa_target <- data.frame(otu_table(phylo_rocky_eDNA))
write.csv(taxa_target, file = "Processed_Data/PRIMER7_format/taxa_target.csv") 

taxonomy_table <- tax_table(phylo_rocky_eDNA)

# CO1
pseq_CO1 <- ps_filter(phylo_rocky_eDNA, primer == "CO1")

meta_target_CO1 <- data.frame(sample_data(pseq_CO1))
write.csv(meta_target_CO1, file = "Processed_Data/PRIMER7_format/meta_target_CO1.csv") 

meta_target_CO1_t <- t(meta_target_CO1)
write.csv(meta_target_CO1_t, file = "Processed_Data/PRIMER7_format/meta_target_CO1_t.csv") 

taxa_target_CO1 <- data.frame(otu_table(pseq_CO1))
write.csv(taxa_target_CO1, file = "Processed_Data/PRIMER7_format/taxa_target_CO1.csv") 

# 18S
pseq_18S <- ps_filter(phylo_rocky_eDNA, primer == "18S")

meta_target_18S <- data.frame(sample_data(pseq_18S))
write.csv(meta_target_18S, file = "Processed_Data/PRIMER7_format/meta_target_18S.csv") 

meta_target_18S_t <- t(meta_target_18S)
write.csv(meta_target_18S_t, file = "Processed_Data/PRIMER7_format/meta_target_18S_t.csv") 

taxa_target_18S <- data.frame(otu_table(pseq_18S))
write.csv(taxa_target_18S, file = "Processed_Data/PRIMER7_format/taxa_target_18S.csv") 
```

We also need some extra meta data for looking at ecological concordance.

```{r}
#| label: import-biogeo-data

biogeo <- read.csv("Input_Data/marclim_categories_tidy.csv")
```

## Map of sampling sites

```{r}
#| label: get-map-elements

#blank map
worldmap = map_data('world')
worldUk <- subset(worldmap, region == c('UK','Ireland'))

#convert to sf object
sf_data <- st_as_sf(worldUk, coords = c("long", "lat"), crs = 4326)  # Specify CRS as WGS84 (EPSG:4326)

# Get countries outline
UK <- rnaturalearth::ne_countries(
  country = c("United Kingdom", "Ireland"),
  scale = "large", returnclass = "sf"
)
```

```{r}
#| label: get-map-lat-longs
 
unique_data <- phylo_rocky_eDNA_long %>% distinct(decimalLongitude, decimalLatitude, verbatimLocality, .keep_all = TRUE)

# Define custom colors for each country
country_colors <- c("Scotland" = "darkseagreen3",
                    "North Wales" = "lightpink2",
                    "Northeast England" = "lightyellow2",
                    "South Wales" = "lightblue2",
                    "Southwest England" = "plum")
```

```{r}
#| label: build-map

site_map <- ggplot() + 
  geom_sf(data = UK ,
          fill = 'grey97', color = 'black') +
  coord_sf(xlim = c(-10, 3), expand = TRUE) +
  scale_x_continuous(sec.axis = dup_axis()) +
  scale_y_continuous(sec.axis = dup_axis()) +
  theme_bw() +
  geom_point(data = unique_data, 
             aes(x = as.numeric(decimalLongitude), 
                 y = as.numeric(decimalLatitude), 
                 fill = country), 
             size = 2, alpha = 1,
             shape = 21, 
             stroke = 0.6) +  # Adjusted size for better visibility
  scale_fill_manual(values = country_colors, guide = "none") +  # Apply custom colors and remove fill legend
  scale_color_discrete(name = "Region") +  # Changed legend name to 'Region'
  ggspatial::annotation_north_arrow(
    location = "bl", which_north = "true",
    style = ggspatial::north_arrow_nautical(
      fill = c("grey40", "white"),
      line_col = "grey20",
      text_family = "ArcherPro Book"
    )
  ) +
  ggspatial::annotation_scale(
    location = "tl",
    bar_cols = c("grey60", "white"),
    text_family = "ArcherPro Book"
  ) +
  theme(legend.position = "left",
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()
  )

ggsave(filename = "Figures/site_map_chpt1.png", plot = site_map,
       device = "png", dpi = 600, units = "mm", height = 200, width = 150)

site_map
```

## Sequence outputs

Explore what our `phyloseq` looks like.

```{r}
#| label: explore-phylo

phylo_rocky_eDNA 
microbiome::summarize_phyloseq(phylo_rocky_eDNA) 
table(meta(phylo_rocky_eDNA)$controlCheck, useNA = "always") #sample type 
table(meta(phylo_rocky_eDNA)$shorePosition, useNA = "always") #sample type
```

We can also look to see how many samples are across different categories like primer, sample type and site.

```{r}
#| label: explore-samples

table(meta(phylo_rocky_eDNA)$primer, useNA = "always") #Primer
table(meta(phylo_rocky_eDNA)$sampleType, useNA = "always") #sample type
table(meta(phylo_rocky_eDNA)$type, useNA = "always") #rockpool vs tidal
table(meta(phylo_rocky_eDNA)$verbatimLocality, useNA = "always") #sites
```

And let's look at the `phyloseq` objects per primer.

```{r}
#| label: primer-data-summaries

microbiome::summarize_phyloseq(phylo_rocky_eDNA_CO1)
microbiome::summarize_phyloseq(phylo_rocky_eDNA_18S)
```

Check how many samples were collected per site. We can see we have an unbalanced sampling effort.

```{r}
#| label: samples-per-site

meta_target$localityID = as.factor(meta_target$localityID)
meta_target$fieldID = as.factor(meta_target$fieldID)
meta_target$sampleType = as.factor(meta_target$sampleType)

# Count unique fieldIDs per site
fieldID_counts <- meta_target %>%
  dplyr::filter(sampleType == "sample") %>%  # Keep only 'sample' rows
  dplyr::group_by(localityID) %>%
  dplyr::summarise(num_fieldIDs = n_distinct(fieldID))
                   
# View results
print(fieldID_counts) # 21 / 32 have full sampling effort
```

## Taxonomy

Let's get some information about taxonomy.

How many functional groups, phyla, families, genera, and taxa are in the data?

```{r}
#| label: explore-taxonomy

length(get_taxa_unique(phylo_rocky_eDNA, "scientificname"))
length(get_taxa_unique(phylo_rocky_eDNA, "valid_name")) #1026
length(get_taxa_unique(phylo_rocky_eDNA, "genus")) #655
length(get_taxa_unique(phylo_rocky_eDNA, "family")) #451
length(get_taxa_unique(phylo_rocky_eDNA, "phylum")) #19
length(get_taxa_unique(phylo_rocky_eDNA, "adult")) #13 unique functional groups
```

Let's explore the taxonomic resolution of our data (i.e. the proportion of each taxonomy ranking achieved).

```{r}
#| label: taxonomic-resolution

is_species_duplicate <- apply(taxonomy_table, 1, function(row) {
  species_name <- tolower(row["scientificname"])  # Convert Species to lower case
  
  # Check if species_name is NA
  if (is.na(species_name)) {
    return(FALSE)
  }
  
  # Compare species name to all higher taxonomic levels ignoring case
  higher_taxa <- tolower(row[1:(which(names(row) == "scientificname") - 1)])  # Convert higher taxa to lower case
  
  # Return TRUE if species_name matches any higher taxon
  return(species_name %in% higher_taxa)
})

taxonomy_table[is_species_duplicate, "scientificname"] <- NA
characterized_taxa <- apply(taxonomy_table, 2, function(x) sum(!is.na(x)))
total_taxa <- nrow(taxonomy_table)
proportion_characterized <- characterized_taxa / total_taxa

print(proportion_characterized)
```

Let's use our rank variable to double check the above function.

```{r}
#| label: taxonomic-resolution-rank

taxonomy_table_summary <- as.data.frame(taxonomy_table) %>% 
  dplyr::group_by(rank) %>% 
  dplyr::summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)

taxonomy_table_summary
```

### Define figure colours

Let's also set-up our colors for all the following plots.

```{r}
#| label: plot_colours

highshore_colour = "#FFD485"
lowshore_colour = "#D44953"
openwater_colour= "#70CBFF"
```

### Venn diagrams

Convert to relative abundances.

```{r}
#| label: relative-abundance-calculaion

pseq.rel <- microbiome::transform(phylo_rocky_eDNA, "compositional")
```

#### Primer crossover

```{r}
#| label: primer-venn-diagram

primers <- unique(as.character(meta(pseq.rel)$primer))

# supplying colors in the order they appear in list_core
mycols <- c("18S"="#96E6B3", "CO1"="#E5A361") 

#all taxa
list_core <- c() # an empty object to store information

for (n in primers){ # for each variable n in primers
  print(paste0("Identifying Core Taxa for ", n))
  
  ps.sub <- subset_samples(pseq.rel, primer == n) # Choose sample from DiseaseState by n
  
  core_m <- get_taxa_unique(prune_taxa(taxa_sums(ps.sub) > 0, ps.sub), "scientificname") 
  
  print(paste0("No. of core taxa in ", n, " : ", length(core_m))) # print core taxa identified in each DiseaseState.
  list_core[[n]] <- core_m # add to a list core taxa for each group.
}

# Specify colors and plot venn
venn_all_primers <- plot(euler(list_core),
     fills = mycols,
     shape = "ellipse",
     quantities = list(type = c("counts", "percent"), cex = 5),
     legend = FALSE,
     labels = list(cex = 6, font = 2),
     lwd = 3)

venn_all_primers
```

#### Shore position crossover

```{r}
#| label: shore-height-venn-diagram

pseq.rel.type <- subset_samples(pseq.rel, shorePosition != "NA")
shore <- unique(as.character(meta(pseq.rel.type)$shorePosition))

# supplying colors in the order they appear in list_core
mycols <- c("High"=highshore_colour,
            "Low"=lowshore_colour,
            "Open Water " = openwater_colour) 

list_core <- c() # an empty object to store information

for (n in shore){ # for each variable n in primers
  print(paste0("Identifying Core Taxa for ", n))
  
  ps.sub <- subset_samples(pseq.rel.type, shorePosition == n) # Choose sample from DiseaseState by n
  
  core_m <- get_taxa_unique(prune_taxa(taxa_sums(ps.sub) > 0, ps.sub), "scientificname") 
  
  print(paste0("No. of core taxa in ", n, " : ", length(core_m))) # print core taxa identified in each DiseaseState.
  list_core[[n]] <- core_m # add to a list core taxa for each group.
}

venn_all_shore <- plot(euler(list_core),
                        fills = mycols,
                        shape = "ellipse",
                        quantities = list(type = c("counts", "percent"), cex = 3),
                        legend = FALSE,
                       labels = list(cex = 3, font = 2),
                       lwd = 3)

venn_all_shore
```

```{r}
#| label: both-venn-diagram

venn_two <- cowplot::plot_grid(
  venn_all_primers,
  venn_all_shore,
  labels = c('a', 'b'),
  label_size = 50,
  nrow = 1
)

venn_two 

ggsave(filename = "Figures/venn_two.png", plot = venn_two,
       device = "png", dpi = 600, units = "mm", height = 550, width = 1250)
```

We can also look at crossover between regions, although the venn diagram is quite complex!

```{r}
#| label: country-venn-diagram

#using another venn function
ggvenn_pq(
  physeq = phylo_rocky_eDNA,
  fact = "country",
  min_nb_seq = 0,
  taxonomic_rank = NULL,
  split_by = NULL,
  add_nb_samples = TRUE,
  add_nb_seq = FALSE,
  rarefy_before_merging = FALSE,
  rarefy_after_merging = FALSE,
  verbose = TRUE,
  type = "nb_taxa",
  na_remove = TRUE
) + ggplot2::scale_fill_distiller(palette = "Blues", direction = 1)

```

#### Summary bar plots

Let's make some simple visualizing of all taxonomic groups together.

```{r}
#| label: taxa-together-kingdom

taxonomy_table <- as.data.frame(taxonomy_table)

# Count the number of species for each phylum
kingdom_species_count<- data.table:::setDT(taxonomy_table)[, .(count = data.table:::uniqueN(valid_AphiaID)), by = kingdom]

kingdom_species_count

# Plot
taxa_groups_bar_kingdom <- ggplot(data = kingdom_species_count, aes(
  x = count,
  y = forcats::fct_rev(reorder(kingdom, kingdom)),
  fill = kingdom
)) +
  geom_bar(stat = "identity") +
  xlab("Number of species") + ylab("Kingdom") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = -0.25,
    hjust = -0.25
  ) +
  scale_fill_manual(
    values = c(
      "skyblue3",
      "lightgreen",
      "grey30",
      "#96E6B3",
      "grey50",
      "#96E6B3",
      "grey50",
      "grey90"
    ),
    name = "Kingdom"
  ) +
  theme_classic()

taxa_groups_bar_kingdom
```

```{r}
#| label: taxa-together-phylum

# Count the number of species for each phylum
phylum_species_count<- data.table:::setDT(taxonomy_table)[, .(count = data.table:::uniqueN(valid_AphiaID)), by = phylum]
phylum_species_count

#add kingdom for fill
kingdom_phyla <- taxonomy_table %>%
  dplyr::select(kingdom, phylum)

phylum_species_count <- na.omit(unique(left_join(phylum_species_count, kingdom_phyla)))

# Plot
taxa_groups_bar_phylum <- ggplot(data = phylum_species_count, aes(
  x = count,
  y = forcats::fct_rev(reorder(phylum, phylum)),
  fill = kingdom
)) +
  geom_bar(stat = "identity") +
  xlab("Number of species") + ylab("Phylum") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust = 0.3,
    hjust = -0.25
  ) +
  scale_fill_manual(values = c(
    "skyblue3",
    "grey50",
    "lightgreen",
    "grey30",
    "#96E6B3",
    "grey20"
  )) +
  theme_classic()

taxa_groups_bar_phylum
```

```{r}
#| label: taxa-together-functional-group

#count number by functional group
function_species_count <- data.table:::setDT(taxonomy_table)[, .(count = data.table:::uniqueN(valid_AphiaID)), by = .(adult_simple, phylum, kingdom)] %>%
  mutate(adult_simple = ifelse(is.na(adult_simple), "Not found", adult_simple))
function_species_count$phylum = as.factor(function_species_count$phylum)

head(function_species_count)

# Group by phylum and adult_simple, and summarize the total count
total_count <- function_species_count %>%
  dplyr::group_by(phylum, adult_simple) %>%  # Grouping by both phylum and simple
  dplyr::summarise(total_count = sum(count, na.rm = TRUE), .groups = 'drop')  # Summarize total count

head(total_count)

# Merge total counts with the original data
function_species_count_reduced <- function_species_count %>% 
  subset(select = c("phylum", "kingdom")) %>%
  distinct()

function_species_count <- left_join(function_species_count_reduced, total_count, by = "phylum")

head(function_species_count)

# order of taxa for the plot
taxa_ordered <- c(
  "Porifera", 
  "Cnidaria", 
  "Ctenophora", 
  "Gastrotricha", 
  "Annelida", 
  "Platyhelminthes",
  "Mollusca", 
  "Arthropoda", 
  "Bryozoa", 
  "Nematoda", 
  "Chaetognatha", 
  "Echinodermata", 
  "Hemichordata", 
  "Chordata", 
  "Phoronida", 
  "Orthonectida",
  "Ochrophyta", 
  "Chlorophyta", 
  "Rhodophyta"
)

# Create the ggplot with custom color mapping
taxa_groups_bar_phylum_functional <- ggplot(data = function_species_count, aes(x = total_count, y = phylum, fill = adult_simple)) +
  geom_bar(stat = "identity") +
  xlab("Number of unique taxa") +
  ylab("Phylum") +
  scale_fill_manual(values = c("Macroalgae" = "#96E6B3",
                               "Macrobenthos" = "#E5A361",
                               "Other" = "grey60",
                               "Not found" = "lightgrey")) +
  labs(fill = "Taxonomic group") +
  scale_y_discrete(limits = taxa_ordered) +
  geom_text(aes(label = total_count), 
            position = position_stack(vjust = 0.5),  # Position text in the center of the bars
            size = 3,  # Adjust size of the text
            color = "black") +  # Change text color for visibility
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 14),  # Make x-axis title slightly bigger
    axis.title.y = element_text(size = 14),  # Make y-axis title slightly bigger
    axis.text.x = element_text(size = 12),   # Make x-axis labels slightly bigger
    axis.text.y = element_text(size = 12),   # Make y-axis labels slightly bigger
    legend.title = element_text(size = 14),  # Make legend title slightly bigger
    legend.text = element_text(size = 12)    # Make legend text slightly bigger
  )


# Print the plot
taxa_groups_bar_phylum_functional

# save
ggsave(filename = "Figures/taxa_groups_bar_phylum_98_curated_functional.png", plot = taxa_groups_bar_phylum_functional,
       device = "png", dpi = 600, units = "mm", height = 200, width = 300)
```

## Diversity analyses

Let's set-up some variables for plotting.

```{r}
#| label: plotting-label-orders

site_order <- c(
  "Scourie",
  "Rispond",
  "Skerray",
  "Murkle Bay",
  "Portskerra",
  "Borwick, Yesnaby",
  "Sannick",
  "Wick",#Scotland
  "Great Orme East",
  "Little Orme",
  "Menai Bridge",
  "Porth Oer",
  "Porth Swtan",
  "Rhosneigr",#North Wales
  "Castlehead Rocks",
  "Filey Brigg",
  "Newton Point",
  "Rumbling Kern",
  "Scalby Mills", #Northumbria
  "Aberystwyth",
  "Neyland",
  "Broad Haven",
  "Skomer Island",
  "West Angle",
  "Monkstone Point",
  "Dale Jetty",
  "Martin's Haven",#South Wales
  "Lizard Point",
  "Looe",
  "Sennen Cove",
  "St Ives",
  "Trevone" #Cornwall
) 

control_order <- c(
  "field-control",
  "lab-negative-control",
  "lab-positive-control",
  "sample",
  "replicate",
  "failed-repeat",
  "stageone-repeat"
)

country_order <- c("Scotland",
                   "Northeast England",
                   "North Wales",
                   "South Wales",
                   "Southwest England")

shore_height_order <- c("High", "Low", "Open Water")
```

### Rarefaction curves

Rarefaction curves are used to check if our read depths look suitable enough to capture all diversity present in the samples. We make a curve for each region and primer.

Scotland

```{r}
#| label: rare-curve-sch

# CO1
SCH_phylo_CO1 <- subset_samples(phylo_rocky_eDNA, country == "Scotland" & primer == 'CO1')
asv_abund_df_SCH_CO1 <- as.data.frame(t(phyloseq::otu_table(SCH_phylo_CO1)))
raremax <- min(rowSums(asv_abund_df_SCH_CO1))
raremax

# Create tidy output
out_SCH_CO1 <- vegan::rarecurve(
  x = asv_abund_df_SCH_CO1,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

# Plot
SCH_CO1_curve <- ggplot(data = out_SCH_CO1, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "#E5A361") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for Scotland CO1 samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()

# 18S
SCH_phylo_18S <- subset_samples(phylo_rocky_eDNA, country == "Scotland" & primer == '18S')
asv_abund_df_SCH_18S <- as.data.frame(t(phyloseq::otu_table(SCH_phylo_18S)))
raremax <- min(rowSums(asv_abund_df_SCH_18S))
raremax

# Create tidy output
out_SCH_18S <- vegan::rarecurve(
  x = asv_abund_df_SCH_18S,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

# Plot
SCH_18S_curve <- ggplot(data = out_SCH_18S, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "green3") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for Scotland 18S samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()
```

South Wales

```{r}
#| label: rare-curve-sw

#CO1
SW_phylo_CO1 <- subset_samples(phylo_rocky_eDNA, country == "South Wales" & primer == 'CO1')
asv_abund_df_SW_CO1 <- as.data.frame(t(phyloseq::otu_table(SW_phylo_CO1)))
raremax <- min(rowSums(asv_abund_df_SW_CO1))
raremax

#Create tidy output
out_SW_CO1 <- vegan::rarecurve(
  x = asv_abund_df_SW_CO1,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
SW_CO1_curve <- ggplot(data = out_SW_CO1, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "#E5A361") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for South Wales CO1 samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()

#18S
SW_phylo_18S <- subset_samples(phylo_rocky_eDNA, country == "South Wales" & primer == '18S')
asv_abund_df_SW_18S <- as.data.frame(t(phyloseq::otu_table(SW_phylo_18S)))
raremax <- min(rowSums(asv_abund_df_SW_18S))
raremax

#Create tidy output
out_SW_18S <- vegan::rarecurve(
  x = asv_abund_df_SW_18S,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
SW_18S_curve <- ggplot(data = out_SW_18S, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "green3") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for South Wales 18S samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()
```

Cornwall

```{r}
#| label: rare-curve-cn

#CO1
CN_phylo_CO1 <- subset_samples(phylo_rocky_eDNA, country == "Southwest England" & primer == 'CO1')
asv_abund_df_CN_CO1 <- as.data.frame(t(phyloseq::otu_table(CN_phylo_CO1)))
raremax <- min(rowSums(asv_abund_df_CN_CO1))
raremax

#Create tidy output
out_CN_CO1 <- vegan::rarecurve(
  x = asv_abund_df_CN_CO1,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
CN_CO1_curve <- ggplot(data = out_CN_CO1, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "#E5A361") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for Cornwall CO1 samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()

#18S
CN_phylo_18S <- subset_samples(phylo_rocky_eDNA, country == "Southwest England" & primer == '18S')
asv_abund_df_CN_18S <- as.data.frame(t(phyloseq::otu_table(CN_phylo_18S)))
raremax <- min(rowSums(asv_abund_df_CN_18S))
raremax

#Create tidy output
out_CN_18S <- vegan::rarecurve(
  x = asv_abund_df_CN_18S,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
CN_18S_curve <- ggplot(data = out_CN_18S, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "green3") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for Cornwall 18S samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()
```

Northeast England

```{r}
#| label: rare-curve-ne

#CO1
NE_phylo_CO1 <- subset_samples(phylo_rocky_eDNA, country == "Northeast England" & primer == 'CO1')
asv_abund_df_NE_CO1 <- as.data.frame(t(phyloseq::otu_table(NE_phylo_CO1)))
raremax <- min(rowSums(asv_abund_df_NE_CO1))
raremax

#Create tidy output
out_NE_CO1 <- vegan::rarecurve(
  x = asv_abund_df_NE_CO1,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
NE_CO1_curve <- ggplot(data = out_NE_CO1, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "#E5A361") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for Northeast CO1 samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()

#18S
NE_phylo_18S <- subset_samples(phylo_rocky_eDNA, country == "Northeast England" & primer == '18S')
asv_abund_df_NE_18S <- as.data.frame(t(phyloseq::otu_table(NE_phylo_18S)))
raremax <- min(rowSums(asv_abund_df_NE_18S))
raremax

#Create tidy output
out_NE_18S <- vegan::rarecurve(
  x = asv_abund_df_NE_18S,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
NE_18S_curve <- ggplot(data = out_NE_18S, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "green3") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for Northeast 18S samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()
```

North Wales

```{r}
#| label: rare-curve-nw

#CO1
NW_phylo_CO1 <- subset_samples(phylo_rocky_eDNA, country == "North Wales" & primer == 'CO1')
asv_abund_df_NW_CO1 <- as.data.frame(t(phyloseq::otu_table(NW_phylo_CO1)))
raremax <- min(rowSums(asv_abund_df_NW_CO1))
raremax

#Create tidy output
out_NW_CO1 <- vegan::rarecurve(
  x = asv_abund_df_NW_CO1,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
NW_CO1_curve <- ggplot(data = out_NW_CO1, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "#E5A361") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for North Wales CO1 samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()

#18S
NW_phylo_18S <- subset_samples(phylo_rocky_eDNA, country == "North Wales" & primer == '18S')
asv_abund_df_NW_18S <- as.data.frame(t(phyloseq::otu_table(NW_phylo_18S)))
raremax <- min(rowSums(asv_abund_df_NW_18S))
raremax

#Create tidy output
out_NW_18S <- vegan::rarecurve(
  x = asv_abund_df_NW_18S,
  step = 20,
  xlab = "Read depth", ylab = "Taxa",
  lwd = 1,
  label = F,
  tidy = T,
  sample = raremax)

#Plot
NW_18S_curve <- ggplot(data = out_NW_18S, mapping = aes(x = Sample, y = Species, group = Site)) +
  geom_line(color = "green3") +
  xlab("Read depth") + ylab("Taxa") +
  #ggtitle("Rarefaction curve for North Wales 18S samples")+
  geom_vline(xintercept =raremax, linetype="dashed") +
  theme_classic()
```

Plot all together.

```{r}
#| label: rare-curve-all

curves_all <- cowplot::plot_grid(NW_18S_curve, 
          NW_CO1_curve,
          SCH_18S_curve,
          SCH_CO1_curve,
          SW_18S_curve,
          SW_CO1_curve,
          CN_18S_curve,
          CN_CO1_curve,
          NE_18S_curve,
          NE_CO1_curve, labels = c('a', 'b', 'c', 'd', 'e', 'f', 'g', 'h', 'i', 'j'), label_size = 12, ncol = 2)

curves_all

ggsave(filename = "Figures/rarefaction_curve_all.png", plot = curves_all,
       device = "png", dpi = 600, units = "mm", height = 450, width = 300)
```

### Alpha diversity

We tested if alpha diversity differed across shore heights through a mixed modelling approach.

```{r}
#| label: calculate-alpha

alpha_df <- phyloseq::estimate_richness(physeq = phylo_rocky_eDNA, measure = c("Observed", "Shannon"))

head(alpha_df)
```

```{r}
#| label: add-data-alpha

# Add diversity metrics to sample data
diversity_meta <- cbind(meta_target, alpha_df)
diversity_meta$shorePosition <- as.factor(diversity_meta$shorePosition)
diversity_meta$verbatimLocality <- as.factor(diversity_meta$verbatimLocality)
diversity_meta <- diversity_meta[!is.na(diversity_meta$shorePosition), ]
diversity_meta <- diversity_meta[!is.na(diversity_meta$country), ]
diversity_meta <- diversity_meta[!is.na(diversity_meta$verbatimLocality), ]
diversity_meta <- diversity_meta[!is.na(diversity_meta$averageTemp), ]
diversity_meta <- diversity_meta[!is.na(diversity_meta$averagePH), ]
```

#### Visualize data

Let's check the distributions of our raw data.

```{r}
#| label: check-alpha-distributions

#response
histogram(sqrt(diversity_meta$Observed)) #normal
histogram((diversity_meta$Shannon)) #fairly normal

#predictors
histogram(diversity_meta$averageTemp) #normal
histogram(diversity_meta$averagePH) #normal
```

```{r}
#| label: mean-temp-over-shore

diversity_meta_summary <- diversity_meta %>% 
  dplyr::group_by(shorePosition) %>% 
  dplyr::summarise(mean = mean(averageTemp),
                   SE = sd(averageTemp) / sqrt(n()))

head(diversity_meta_summary) 

ggplot(diversity_meta, aes(x = shorePosition, y = averageTemp)) + 
  geom_boxplot() +
  theme_classic()
```

```{r}
#| label: visual-alpha-temp

# richness vs temp
diversity_temp_scatter <- ggplot(diversity_meta,
                                 aes(x = averageTemp, y = Observed, colour = shorePosition)) +
  geom_point() +
  theme_classic()

diversity_temp_scatter

# shannon vs temp
diversity_temp_scatter <- ggplot(diversity_meta, aes(x = averageTemp, y = Shannon)) +
  geom_point() +
  theme_classic()
```

```{r}
#| label: visual-alpha-ph

# richness vs pH
diversity_pH_scatter <- ggplot(diversity_meta, aes(x = averagePH, y = Observed)) +
  geom_point() +
  theme_classic()

diversity_pH_scatter #no pattern observed

# shannon vs pH
diversity_pH_scatter <- ggplot(diversity_meta, aes(x = averagePH, y = Shannon)) +
  geom_point() +
  theme_classic()

diversity_pH_scatter #no pattern observed
```

#### Mixed models

```{r}
#| label: richness-mixed-model

# construct model
richness_mod <- glmer(
  Observed ~ country + shorePosition + averageTemp + averagePH +
    (1 | country:verbatimLocality),
  data = diversity_meta,
  family = poisson,
  control = glmerControl(optimizer = "bobyqa")
)

#outputs
summary(richness_mod)
anova(richness_mod)
Anova(richness_mod, type = "II")

# get p-values
beta_hat <- fixef(richness_mod)  # Estimated coefficients
se_beta <- sqrt(diag(vcov(richness_mod)))  # Standard errors of the coefficients
t_values <- beta_hat / se_beta # Calculate Wald t-values manually
p_values <- 2 * (1 - pnorm(abs(t_values)))  # Two-tailed test
p_values
```

```{r}
#| label: richness-mixed-model-assumptions

plot(richness_mod) #good
ranef(richness_mod)
vif(richness_mod)
qqnorm(resid(richness_mod))
qqline(resid(richness_mod)) #tail but not too bad
```

```{r}
#| label: richness-mixed-model-cal-variation

# random variance
123.0/(211.4   + 123.0)*100 #36.78% for the nested random component
```

```{r}
#| label: richness-random-se

sqrt(diag(vcov(richness_mod, full = TRUE)))
vv <- vcov(richness_mod, full = TRUE)
colnames(vv)
sqrt(diag(vv)[grepl("^cov", colnames(vv))]) # pick out the random components
```

Let's do the same again for Shannon.

```{r}
#| label: shannon-mixed-model

# build model
#diversity_meta$Shannon_adj <- diversity_meta$Shannon + 0.001

shannon_mod <- lmer(
  Shannon ~ country + shorePosition + averageTemp + averagePH +
    (1 | country:verbatimLocality),
  data = diversity_meta
)

# outputs
summary(shannon_mod)
anova(shannon_mod)
Anova(shannon_mod, type = "II")

# get p-values
beta_hat <- fixef(shannon_mod)  # Estimated coefficients
se_beta <- sqrt(diag(vcov(shannon_mod)))  # Standard errors of the coefficients
t_values <- beta_hat / se_beta # Calculate Wald t-values manually
p_values <- 2 * (1 - pnorm(abs(t_values)))  # Two-tailed test
p_values

# get df
summary(shannon_mod, ddf = "Kenward-Roger") 
anova(shannon_mod, ddf = "Kenward-Roger")
```

```{r}
#| label: shannon-mixed-model-assumptions

plot(shannon_mod) #good
ranef(shannon_mod)
vif(shannon_mod)
qqnorm(resid(shannon_mod))
qqline(resid(shannon_mod)) #tail but not too bad
```

```{r}
#| label: shannon-mixed-model-cal-variation

# check random variance
0.06655 /(0.06655   + 0.44087 )*100 #13.11537% for the nested random component
```

```{r}
#| label: shannon-random-se

sqrt(diag(vcov(shannon_mod, full = TRUE)))
vv <- vcov(shannon_mod, full = TRUE)
colnames(vv)
sqrt(diag(vv)[grepl("^cov", colnames(vv))]) # pick out the random components
```

#### Planned contrasts

```{r}
#| label: set-contrast-levels

Open = c(0, 0, 1)
Low = c(0, 1, 0)
High= c(1, 0, 0)
Low_and_high = (Low + High)/2
```

```{r}
#| label: contrast-richness

emms_richness <- emmeans::emmeans(richness_mod, ~ shorePosition)
emms_richness

contrast_richness_1 <- contrast(emms_richness, method = list("High - Low" = High - Low))
contrast_richness_2 <- contrast(emms_richness, method = list("Open - Other" = Open - Low_and_high))

contrast_richness_1 
contrast_richness_2

# Get confidence intervals for richness contrasts
confint(contrast_richness_1)
confint(contrast_richness_2)
```

```{r}
#| label: contrast-shannon

emms_shannon <- emmeans::emmeans(shannon_mod, ~ shorePosition)
emms_shannon

contrast_shannon_1 <- contrast(emms_shannon, method = list("High - Low" = High - Low))
contrast_shannon_2 <- contrast(emms_shannon, method = list("Open - Other" = Open - Low_and_high))

contrast_shannon_1
contrast_shannon_2

# Get confidence intervals for Shannon contrasts
confint(contrast_shannon_1)
confint(contrast_shannon_2)
```

#### Box and scatter plots

```{r}
#| label: richness-shore-height-boxplot

rich_shore <- ggplot(aes(shorePosition, Observed, fill = shorePosition), data = diversity_meta) +
  geom_boxplot() +
  #facet_wrap(~ country) + # create a facet for each mountain range
  xlab("Shore height") +
  ylab("Richness") +
  scale_fill_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    'Open Water' = openwater_colour
  )) +
  #geom_signif(comparisons = list(c("High", "Low")),
  #map_signif_level = TRUE, , tip_length = 0,
  #annotation = "***")+
  #geom_signif(comparisons = list(c("High", "Open Water")),
  #map_signif_level = TRUE, tip_length = 0, y_position = 124,
  #annotation = "***") +
  ylim(0, 140) +
  theme_classic() +
  theme(legend.position = "none")

rich_shore 
```

```{r}
#| label: shannon-shore-height-boxplot

shannon_shore <- ggplot(aes(shorePosition, Shannon, fill = shorePosition), data = diversity_meta) +
  geom_boxplot() +
  #facet_wrap(~ country) + # create a facet for each mountain range
  xlab("Shore height") +
  ylab("Shannon Diversity Index") +
  #geom_signif(comparisons = list(c("High", "Low")),
  #map_signif_level = TRUE, , tip_length = 0,
  #annotation = "NS")+
  #eom_signif(comparisons = list(c("High", "Open Water")),
  #map_signif_level = TRUE, tip_length = 0, y_position = 4.05,
  #annotation = "***") +
  ylim(0, 4.5) +
  scale_fill_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    'Open Water' = openwater_colour
  )) +
  theme_classic() +
  theme(legend.position = "none")

shannon_shore
```

```{r}
#| label: richness-temp-scatter-pred

# Extract the prediction data frame
pred.temp.rich <- ggpredict(richness_mod, terms = c("averageTemp [all]", "shorePosition"), bias_correction = TRUE)

#plot(pred.temp.rich, show_data = TRUE, jitter = TRUE)

# Plot
richness_temp <- ggplot() +
  # Add observed data points
  geom_point(
    data = diversity_meta,
    aes(x = averageTemp, y = Observed, colour = shorePosition),
    size = 1.8,
    alpha = 0.5
  ) +
  
  # Add predicted lines with color mapping by shorePosition
  geom_line(data = pred.temp.rich,
            aes(x = x, y = predicted, colour = group),
            linewidth = 1.2) +
  
  # Add confidence intervals (shading) by shorePosition
  geom_ribbon(
    data = pred.temp.rich,
    aes(
      x = x,
      ymin = conf.low,
      ymax = conf.high,
      fill = group
    ),
    alpha = 0.5
  ) +
  
  # Set manual colors for shorePosition
  scale_colour_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  scale_fill_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  # Labels and theme
  xlab("Temperature (°C)") +
  ylab("Richness") +
  xlim(10, 26) +
  labs(colour = "Shore Position", fill = "Shore Position") +
  theme_classic()+
  theme(legend.position = "none")

richness_temp
```

```{r}
#| label: richness-ph-scatter-pred

# Extract the prediction data frame
pred.ph.rich <- ggpredict(richness_mod, terms = c("averagePH [all]", "shorePosition"), bias_correction = TRUE)

#plot(pred.ph.rich, show_data = TRUE, jitter = TRUE)

# Plot
richness_ph <- ggplot() +
  # Add observed data points
  geom_point(
    data = diversity_meta,
    aes(x = averagePH, y = Observed, colour = shorePosition),
    size = 1.8,
    alpha = 0.5
  ) +
  
  # Add predicted lines with color mapping by shorePosition
  geom_line(data = pred.ph.rich,
            aes(x = x, y = predicted, colour = group),
            linewidth = 1.2) +
  
  # Add confidence intervals (shading) by shorePosition
  geom_ribbon(
    data = pred.ph.rich,
    aes(
      x = x,
      ymin = conf.low,
      ymax = conf.high,
      fill = group
    ),
    alpha = 0.5
  ) +
  
  # Set manual colors for shorePosition
  scale_colour_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  scale_fill_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  # Labels and theme
  xlab("pH") +
  ylab("Richness") +
  #xlim(10, 26) +
  labs(colour = "Shore Position", fill = "Shore Position") +
  theme_classic()+
  theme(legend.position = "none")

richness_ph
```

```{r}
#| label: shannon-temp-scatter-pred

# Extract the prediction data frame
pred.temp.shannon <- ggpredict(shannon_mod, terms = c("averageTemp", "shorePosition"))

# plot
shannon_temp <- ggplot() +
  # Add observed data points
  geom_point(
    data = diversity_meta,
    aes(x = averageTemp, y = Shannon, colour = shorePosition),
    size = 1.8,
    alpha = 0.5
  ) +
  
  # Add predicted lines with color mapping by shorePosition
  geom_line(data = pred.temp.shannon,
            aes(x = x, y = predicted, colour = group),
            size = 1.2) +
  
  # Add confidence intervals (shading) by shorePosition
  geom_ribbon(
    data = pred.temp.shannon,
    aes(
      x = x,
      ymin = predicted - std.error,
      ymax = predicted + std.error,
      fill = group
    ),
    alpha = 0.3
  ) +
  
  # Set manual colors for shorePosition
  scale_colour_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  scale_fill_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  # Labels and theme
  xlab("Temperature (°C)") +
  ylab("Shannon Diversity Index") +
  xlim(10, 25) +
  labs(colour = "Shore Position", fill = "Shore Position") +
  theme_classic() +
  theme(legend.position = "none")

shannon_temp 
```

```{r}
#| label: shannon-ph-scatter-pred

# Extract the prediction data frame
pred.ph.shannon <- ggpredict(shannon_mod, terms = c("averagePH", "shorePosition"))

# plot
shannon_ph<- ggplot() +
  # Add observed data points
  geom_point(
    data = diversity_meta,
    aes(x = averagePH, y = Shannon, colour = shorePosition),
    size = 1.8,
    alpha = 0.5
  ) +
  
  # Add predicted lines with color mapping by shorePosition
  geom_line(data = pred.ph.shannon,
            aes(x = x, y = predicted, colour = group),
            size = 1.2) +
  
  # Add confidence intervals (shading) by shorePosition
  geom_ribbon(
    data = pred.ph.shannon,
    aes(
      x = x,
      ymin = predicted - std.error,
      ymax = predicted + std.error,
      fill = group
    ),
    alpha = 0.3
  ) +
  
  # Set manual colors for shorePosition
  scale_colour_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  scale_fill_manual(values = c(
    "Low" = lowshore_colour,
    "High" = highshore_colour,
    "Open Water" = openwater_colour
  )) +
  
  # Labels and theme
  xlab("pH") +
  ylab("Shannon Diversity Index") +
  #xlim(10, 25) +
  labs(colour = "Shore Position", fill = "Shore Position") +
  theme_classic() +
  theme(legend.position = "none")

shannon_ph
```

```{r}
#| label: alpha-figure-final-plot

alpha_plot_all <- plot_grid(rich_shore, richness_temp, shannon_shore, shannon_temp, labels = c("a", "b", "c", "d"), ncol = 2, align = "v")

#save
ggsave(filename = "Figures/alpha_plot_all.png", plot = alpha_plot_all ,
       device = "png", dpi = 600, units = "mm", height = 200, width = 230)

alpha_plot_all
```

```{r}
#| label: alpha-figure-final-plot-ph

alpha_plot_all2 <- plot_grid(rich_shore, shannon_shore, richness_temp, shannon_temp, richness_ph, shannon_ph, labels = c("a", "b", "c", "d", "e", "f", "g"), ncol = 2, align = "v")

#save
ggsave(filename = "Figures/alpha_plot_all_2.png", plot = alpha_plot_all2 ,
       device = "png", dpi = 600, units = "mm", height = 300, width = 230)

alpha_plot_all2
```

#### Split by algal and inverts

We can also look at diversity across invertebrates and algal separately.

```{r}
#| label: define-inverts-algae

invert_phyla <- c("Acoela",
                  "Amphioxus",
                  "Annelida",
                  "Arthropoda",
                  "Brachiopoda",
                  "Bryozoa",
                  "Cephalochordata",
                  "Chaetognatha",
                  "Chordata",
                  "Cnidaria",
                  "Crustacea",
                  "Echiura",
                  "Echinoderm",
                  "Echinodermata",
                  "Gastrotricha",
                  "Gnathostomulids",
                  "Hemichordata",
                  "Kamptozoa",
                  "Kinorhyncha",
                  "Loricifera",
                  "Mollusca",
                  "Monogenea",
                  "Myzostomida",
                  "Nematoda",
                  "Nemertinea",
                  "Orthonectida",
                  "Phaeophyta",
                  "Phoronida",
                  "Placozoa",
                  "Platyhelminthes",
                  "Porifera",
                  "Priapulida",
                  "Pycnogonida",
                  "Sipunculida",
                  "Thallophyta",
                  "Tunicata",
                  "Xenoturbella",
                  "Xiphosura") #38
                  
algae_phyla <- c("Chlorophyta",
                 "Ctenophora",
                 "Dinoflagellata",
                 "Euglenophyta",
                 "Gyrista",
                 "Ochrophyta",
                 "Phaeophyta",
                 "Rhodophyta",
                 "Thallophyta")
```

```{r}
#| label: filter-phylo-inverts-algae

physeq_invert <- subset_taxa(phylo_rocky_eDNA, phylum %in% invert_phyla)
physeq_invert

physeq_algae <- subset_taxa(phylo_rocky_eDNA, phylum %in% algae_phyla)
physeq_algae

physeq_invert_noNA <- ps_filter(physeq_invert, country != "NA")
physeq_algae_noNA <- ps_filter(physeq_algae, country != "NA")
```

```{r}
#| label: wrangle-for-comparison-plot-algae-inverts

richness_invert <- estimate_richness(physeq_invert_noNA, measures = c("Observed", "Shannon")) %>% mutate(physeq_object = "Invertebrates")

richness_algae <- estimate_richness(physeq_algae_noNA, measures = c("Observed", "Shannon")) %>% mutate(physeq_object = "Algae")

combined_richness <- bind_rows(richness_invert, richness_algae)

# Add sample metadata (shore position) from your phyloseq objects
combined_richness$shorePosition <- c(
  sample_data(physeq_invert_noNA)$shorePosition,
  sample_data(physeq_algae_noNA)$shorePosition
)

# Add sample metadata (region) from your phyloseq objects
combined_richness$region <- c(sample_data(physeq_invert_noNA)$country,
                              sample_data(physeq_algae_noNA)$country)

combined_richness <- tibble::rownames_to_column(combined_richness, "sample")

# Reshape the data into a long format
long_richness <- combined_richness %>%
  pivot_longer(
    cols = c(Observed, Shannon),
    names_to = "Diversity_Metric",
    values_to = "Value"
  )
```

```{r}
#| label: plot-algae-invert-richness-region

split_diversity_region <- ggplot(long_richness, aes(x = region, y = Value, colour = physeq_object)) +
  geom_boxplot() +
  scale_x_discrete(limits = country_order) +
  facet_wrap( ~ Diversity_Metric, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "Algae" = "green4",
      "Invertebrates" = "orange2",
      "Total" = "grey70"
    ),
    name = "Taxonomic group"
  ) +
  xlab("Region") +
  ylab("Diversity Index Value") +
  theme_bw() +
  guides(fill = guide_legend(title = "Taxonomic group")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

split_diversity_region 
```

```{r}
#| label: plot-algae-invert-richness-shoreheight

split_diversity_type <- ggplot(long_richness,
                               aes(x = shorePosition, y = Value, colour = physeq_object)) +
  geom_boxplot() +
  facet_wrap( ~ Diversity_Metric, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "Algae" = "green4",
      "Invertebrates" = "orange2",
      "Total" = "grey70"
    ),
    name = "Taxonomic group"
  ) +
  scale_x_discrete(breaks = shore_height_order) +
  xlab("Shore position") +
  ylab("Diversity Index Value") +
  theme_bw() +
  guides(fill = guide_legend(title = "Taxonomic group"))

split_diversity_type
```

Join for export.

```{r}
#| label: plot-algae-invert-richness-both

split_both <- cowplot::plot_grid(
  split_diversity_region,
  split_diversity_type,
  labels = c("a", "b"),
  align = "v",
  ncol = 1,
  rel_heights = c(1.3, 1)
)

split_both

ggsave(filename = "Figures/alpha_both_split.png", plot = split_both,
       device = "png", dpi = 600, units = "mm", height = 300, width = 300)
```

### Beta diversity

While our stats are performed in PRIMER7, we create our NMDS and dissimilarity plots here.

```{r}
#| label: beta-data-set-up

# filter out any NAs for variables of interest
physeq_beta <- ps_filter(phylo_rocky_eDNA, shorePosition != "NA")
physeq_beta <- ps_filter(physeq_beta, country != "NA")
physeq_beta <- prune_samples(sample_sums(physeq_beta) > 0, physeq_beta)

# convert to P/A
physeq_beta_PA <- microbiome::transform(physeq_beta,"pa")

# split by gene
pseq_CO1_pruned <- ps_filter(physeq_beta_PA, primer == "CO1")
pseq_18S_pruned <- ps_filter(physeq_beta_PA, primer == "18S")
```

#### NMDS plots

```{r}
#| label: calculate-nmds-jaccard-co1

ord.nmds.jaccard.CO1 <- ordinate(pseq_CO1_pruned, method = "NMDS", distance =
                                   "jaccard", k = 3)

stress_value_jaccard_CO1 <- ord.nmds.jaccard.CO1$stress
```

```{r}
#| label: calculate-nmds-jaccard-18s

ord.nmds.jaccard.18S <- ordinate(pseq_18S_pruned, method = "NMDS", distance =
                                   "jaccard", k = 3)
stress_value_jaccard_18S <- ord.nmds.jaccard.18S$stress
```

```{r}
#| label: nmds-region-jaccard-co1

NMDS_region_CO1_jaccard <- plot_ordination(pseq_CO1_pruned, ord.nmds.jaccard.CO1, color="country")+
  scale_colour_manual(values = c("Scotland" = "darkseagreen",
                                 "North Wales" = "lightpink3",
                                 "Northeast England" = "yellow3",
                                 "South Wales" = "lightblue",
                                 "Southwest England" = "mediumpurple")) +
  theme_classic()+ 
  #theme(legend.position = "none")+ 
  annotate("text", x = Inf, y = Inf, label = paste("Stress =", round(stress_value_jaccard_CO1, 3)), 
           size = 4, hjust = 1.1, vjust = 1.5)+
  #stat_ellipse(aes(group = interaction(country, shorePosition)), type = "t", level = 0.95, linetype = 1, size = 0.5)+
  stat_ellipse(aes(group = country), type = "t", level = 0.95, linetype = 1, size = 0.5)+
  #annotate("text", x = -0.6, y = 0.6, label = paste("CO1"), 
           #size = 4)+
  labs(color = "Region") +
  theme(legend.position = "bottom") +
  theme(legend.position = "none")+ 
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) 

NMDS_region_CO1_jaccard
```

```{r}
#| label: nmds-region-jaccard-18S

NMDS_region_18S_jaccard <- plot_ordination(pseq_18S_pruned, ord.nmds.jaccard.18S, color="country")+
  scale_colour_manual(values = c("Scotland" = "darkseagreen",
                                 "North Wales" = "lightpink3",
                                 "Northeast England" = "yellow3",
                                 "South Wales" = "lightblue",
                                 "Southwest England" = "mediumpurple")) +
  theme_classic()+
  theme(legend.position = "none")+ 
  annotate("text", x = Inf, y = Inf, label = paste("Stress =", round(stress_value_jaccard_18S, 3)), 
           size = 4, hjust = 1.1, vjust = 1.5)+
  stat_ellipse(aes(group = country), type = "t", linetype = 1, size = 0.5)+
  #annotate("text", x = -1.1, y = 0.9, label = paste("18S"), 
           #size = 4)+
  xlim(-1.2,1.8)+ #to remove weird outlier
  labs(color = "Region") 


NMDS_region_18S_jaccard
```

```{r}
#| label: nmds-shore-height-jaccard-co1

NMDS_type_CO1_jaccard<- plot_ordination(pseq_CO1_pruned, ord.nmds.jaccard.CO1, color="shorePosition") +
  scale_colour_manual(values = c("Low" = lowshore_colour, 
                               "High" = highshore_colour,
                               'Open Water' = openwater_colour)) +
  theme_classic()+
  #theme(legend.position = "none")+ 
  annotate("text", x = Inf, y = Inf, label = paste("Stress =", round(stress_value_jaccard_CO1, 3)), 
           size = 4, hjust = 1.1, vjust = 1.5)+
  stat_ellipse(aes(group = shorePosition), type = "t", linetype = 1, size = 0.5)+
  #annotate("text", x = -0.6, y = 0.6, label = paste("CO1"), 
           #size = 4)+
  labs(color = "Shore position") +
  theme(legend.position = "none") +
  guides(colour = guide_legend(nrow = 1, byrow = TRUE)) 

NMDS_type_CO1_jaccard
```

```{r}
#| label: nmds-shore-height-jaccard-18s

NMDS_type_18S_jaccard <- plot_ordination(pseq_18S_pruned, ord.nmds.jaccard.18S, color="shorePosition") +
  scale_colour_manual(values = c("Low" = lowshore_colour, 
                                 "High" = highshore_colour,
                                 'Open Water' = openwater_colour)) +
  theme_classic()+
  theme(legend.position = "none")+ 
  annotate("text", x = Inf, y = Inf, label = paste("Stress =", round(stress_value_jaccard_18S, 3)), 
           size = 4, hjust = 1.1, vjust = 1.5)+
  stat_ellipse(aes(group = shorePosition), type = "t", linetype = 1, size = 0.5)+
  #annotate("text", x = -1.1, y = 1, label = paste("18S"), 
           #size = 4)+
  xlim(-1.1, 1.8)+
  labs(color = "Shore position")

NMDS_type_18S_jaccard
```

```{r}
#| label: nmds-all-plot-export

NMDS_jaccard_all<- cowplot::plot_grid(NMDS_region_18S_jaccard, NMDS_type_18S_jaccard,NMDS_region_CO1_jaccard, NMDS_type_CO1_jaccard, labels = c("a", "b", "c", "d"), ncol = 2, align = "hv")

NMDS_jaccard_all

ggsave("Figures/NMDS_jaccard_all.png", NMDS_jaccard_all, units = "mm", width = 240, height = 240)
```

#### Check for homogeneity of dispersion

```{r}
#| label: test-homogeneity-dispersion

#CO1 - region
jaccard_dist_CO1 <- distance(pseq_CO1_pruned, method = "jaccard")
region <- sample_data(pseq_CO1_pruned)$country
bd_CO1_r <- betadisper(d = jaccard_dist_CO1, group = region, type = "centroid")
permutest(bd_CO1_r) #signficant

#CO1 - shore height
shore_height <- sample_data(pseq_CO1_pruned)$shorePosition
bd_CO1_sh <- betadisper(d =jaccard_dist_CO1, group = shore_height, type = "centroid")
permutest(bd_CO1_sh) #signficant

#CO1 - sites
sites <- sample_data(pseq_CO1_pruned)$verbatimLocality 
bd_CO1_s <- betadisper(d =jaccard_dist_CO1, group = sites, type = "centroid")
permutest(bd_CO1_s) #signficant

#18S - region
jaccard_dist_18S <- distance(pseq_18S_pruned, method = "jaccard")
region <- sample_data(pseq_18S_pruned)$country
bd_18S_r <- betadisper(d = jaccard_dist_18S, group = region, type = "centroid")
permutest(bd_18S_r) #signficant

#18S - shore height
shore_height <- sample_data(pseq_18S_pruned)$shorePosition
bd_18S_sh <- betadisper(d =jaccard_dist_18S, group = shore_height, type = "centroid")
permutest(bd_18S_sh) #signficant

#18S - sites
sites <- sample_data(pseq_18S_pruned)$verbatimLocality 
bd_18S_s <- betadisper(d =jaccard_dist_18S, group = sites, type = "centroid")
permutest(bd_18S_s) #signficant
```

#### Pairwise dissimilarity boxplots

```{r}
#| label: jaccard-matrix-co1

jaccard_dist_CO1 <- phyloseq::distance(pseq_CO1_pruned, method="jaccard")

# format matrix to long
jaccard_dist_CO1_mat <- as.matrix(jaccard_dist_CO1)
longData_CO1<-melt(jaccard_dist_CO1_mat)
longData_CO1<-longData_CO1[longData_CO1$value!=0,]
```

```{r}
#| label: subset-meta-co1

metadf <- data.frame(sample_data(pseq_CO1_pruned))
metadf_shore <- metadf %>% subset(select = c("fullID", "verbatimLocality", "shorePosition", "country"))
```

```{r}
#| label: format-data-for-comparisons-CO1

# Join longData_CO1 with metadf_shore to get Var1_position
longData_CO1 <- longData_CO1 %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var1_position = shorePosition),
    by = c("Var1" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_CO1 <- longData_CO1 %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var2_position = shorePosition),
    by = c("Var2" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_CO1 <- longData_CO1 %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var1_site = verbatimLocality),
    by = c("Var1" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_CO1 <- longData_CO1 %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var2_site = verbatimLocality),
    by = c("Var2" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_CO1 <- longData_CO1 %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var1_country = country),
    by = c("Var1" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_CO1 <- longData_CO1 %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var2_country = country),
    by = c("Var2" = "fullID")
  )

# Create the comparison variable
longData_CO1 <- longData_CO1 %>%
  mutate(
    # Create a consistent comparison column by sorting positions
    comparison = pmap_chr(list(Var1_position, Var2_position), function(v1, v2) {
      if (v1 == v2) {
        "Within shore position"
      } else {
        sorted <- sort(c(v1, v2)) # Sort the positions
        paste(sorted, collapse = "") # Concatenate the sorted positions
      }
    }),
    sameSite = Var1_site == Var2_site,
    samePosition = Var1_position == Var2_position
  )

longData_CO1$comparison[longData_CO1$comparison == "HighOpen Water"] = "HighOpenWater"
longData_CO1$comparison[longData_CO1$comparison == "LowOpen Water"] = "LowOpenWater"
longData_CO1_samesites <- longData_CO1 %>% filter(sameSite == TRUE)
```

```{r}
#| label: summarise-data-for-comparisons-CO1

# Plotting the boxplot using ggplot2
standard_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

which(longData_CO1_samesites$Var1_site != longData_CO1_samesites$Var2_site)

# Calculate the averages for each group site
averages <- longData_CO1_samesites %>%
  dplyr::group_by(comparison, Var1_site) %>%
  dplyr::summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop',
                   standard_error = standard_error(value))

averages

# Calculate the averages for each group region
averages_regions <- longData_CO1_samesites %>%
  dplyr::group_by(comparison, Var1_country) %>%
  dplyr::summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop',
                   standard_error = standard_error(value))

averages_regions

# Calculate the averages for each group region
averages_compare_CO1 <- longData_CO1_samesites %>%
  dplyr::group_by(comparison) %>%
  dplyr::summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop',
                   standard_error = standard_error(value))
averages_compare_CO1

# Define the four unique comparisons you're interested in
required_comparisons <- c("HighOpenWater",
                          "HighLow",
                          "LowOpenWater",
                          "Within shore position")

# Filter the sites that have all required comparisons
filtered_sites <- longData_CO1_samesites #%>%
  #dplyr::group_by(Var1_site) %>%
  #dplyr::filter(all(required_comparisons %in% comparison)) %>%
  #dplyr::ungroup()
```

```{r}
#| label: label-variables-diss-comparisons

comparison_order <- c("High vs Open Water",
                      "High vs Low",
                      "Low vs Open Water",
                      "Same shore position")

# needed if using full sampling effort sites only
site_order_reduced <- c(
  "Scourie",
  "Skerray",
  "Murkle Bay",
  "Wick", #Scotland
  "Porth Swtan",
  "Rhosneigr",#North Wales
  "Castlehead Rocks",
  "Filey Brigg",
  "Newton Point",
  "Rumbling Kern",
  "Scalby Mills",#Northumbria
  "Aberystwyth",
  "Broad Haven",
  "West Angle",#South Wales
  "Lizard Point",
  "Looe",
  "Sennen Cove",
  "St Ives",
  "Trevone" #Cornwall
) 

# Reorder the factor levels for the comparison variable
filtered_sites$comparison <- factor(
  filtered_sites$comparison,
  levels = c(
    "HighOpenWater",
    "HighLow",
    "LowOpenWater",
    "Within shore position"
  )
)
```

```{r}
#| label: plot-jaccard-comparisons-regions-co1

compare_jaccard_regions <- ggplot(filtered_sites, aes(x = Var1_country, y = value, fill = comparison, colour = comparison)) +
  geom_boxplot(alpha = 0.5) +
  #geom_point(position = position_jitterdodge(dodge.width = .5), alpha = 0.2) +
  #geom_text(data = averages, aes(x = Var1_site, y = mean_value, label = round(mean_value, 2)),
  #vjust = -1.8, color = "black", size = 3.5, position = position_dodge(width = 0.75)) +
  labs(x = "Region",
       y = "Jaccard Dissimilarity (COI)") +
  scale_fill_manual(values = c("HighLow" = "#7DCE82", 
                               "Within shore position" = "grey50",
                               "LowOpenWater" = "#FF8811",
                               "HighOpenWater" = "#5998C5"), 
                    labels = c("HighLow" = "High vs Low", 
                               "HighOpenWater" = "High vs Open Water",
                               "LowOpenWater" = "Low vs Open Water",
                               "Within shore position" = "Same shore height"),
                    name = "Shore Height Comparison") +
  scale_colour_manual(values = c("HighLow" = "green4", 
                                 "Within shore position" = "grey30",
                                 "LowOpenWater" = "orange3",
                                 "HighOpenWater" = "skyblue4"),
                      labels = c("HighLow" = "High vs Low", 
                                 "HighOpenWater" = "High vs Open Water",
                                 "LowOpenWater" = "Low vs Open Water",
                                 "Within shore position" = "Same shore height"),
                      name = "Shore Height Comparison")+
  theme_classic() +
  scale_x_discrete(limits = country_order) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'none') + # Remove x-axis title)
scale_y_continuous(labels = scales::number_format(accuracy = 0.1))

compare_jaccard_regions 
```

Same again for 18S.

```{r}
#| label: jaccard-matrix-18S

#Produce distance matrix
jaccard_dist_18S <- phyloseq::distance(pseq_18S_pruned, method="jaccard")

#subset meta data for shore heights
metadf <- data.frame(sample_data(pseq_18S_pruned))
metadf_shore <- metadf %>% subset(select = c("fullID", "verbatimLocality", "shorePosition", "country"))

# format matrix to long
jaccard_dist_18S_mat <- as.matrix(jaccard_dist_18S)
longData_18S<-melt(jaccard_dist_18S_mat)
longData_18S<-longData_18S[longData_18S$value!=0,]
```

```{r}
#| label: format-data-for-comparisons-18s

# Join longData_CO1 with metadf_shore to get Var1_position
longData_18S <- longData_18S %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var1_position = shorePosition),
    by = c("Var1" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_18S <- longData_18S %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var2_position = shorePosition),
    by = c("Var2" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_18S <- longData_18S %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var1_site = verbatimLocality),
    by = c("Var1" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_18S <- longData_18S %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var2_site = verbatimLocality),
    by = c("Var2" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_18S <- longData_18S %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var1_country = country),
    by = c("Var1" = "fullID")
  )

# Join longData_CO1 with metadf_shore again to get Var2_position
longData_18S <- longData_18S %>%
  left_join(
    metadf_shore %>% 
      dplyr::select(fullID, Var2_country = country),
    by = c("Var2" = "fullID")
  )

# Create the comparison variable
longData_18S <- longData_18S %>%
  mutate(
    # Create a consistent comparison column by sorting positions
    comparison = pmap_chr(list(Var1_position, Var2_position), function(v1, v2) {
      if (v1 == v2) {
        "Within shore position"
      } else {
        sorted <- sort(c(v1, v2)) # Sort the positions
        paste(sorted, collapse = "") # Concatenate the sorted positions
      }
    }),
    sameSite = Var1_site == Var2_site,
    samePosition = Var1_position == Var2_position
  )

longData_18S$comparison[longData_18S$comparison == "HighOpen Water"] = "HighOpenWater"
longData_18S$comparison[longData_18S$comparison == "LowOpen Water"] = "LowOpenWater"

longData_18S_samesites <- longData_18S %>% filter(sameSite == TRUE)

which(longData_18S_samesites$Var1_site != longData_18S_samesites$Var2_site)
```

```{r}
#| label: summarise-data-for-comparisons-18s

# Calculate the averages for each group
averages <- longData_18S_samesites %>%
  dplyr::group_by(comparison, Var1_site) %>%
  dplyr::summarise(
    mean_value = mean(value, na.rm = TRUE),
    .groups = 'drop',
    standard_error = standard_error(value)
  )

averages

# Calculate the averages for each group
averages_regions <- longData_18S_samesites %>%
  dplyr::group_by(comparison, Var1_country) %>%
  dplyr::summarise(
    mean_value = mean(value, na.rm = TRUE),
    .groups = 'drop',
    standard_error = standard_error(value)
  )

averages_regions

averages_compare_18S <- longData_18S_samesites %>%
  dplyr::group_by(comparison) %>%
  dplyr::summarise(
    mean_value = mean(value, na.rm = TRUE),
    .groups = 'drop',
    standard_error = standard_error(value)
  )

averages_compare_18S

# Filter the sites that have all required comparisons
filtered_sites_18S <- longData_18S_samesites #%>%
  #group_by(Var1_site) %>%
  #dplyr::filter(all(required_comparisons %in% comparison)) %>%
  #ungroup()

# Reorder the factor levels for the comparison variable
filtered_sites_18S$comparison <- factor(
  filtered_sites_18S$comparison,
  levels = c(
    "HighOpenWater",
    "HighLow",
    "LowOpenWater",
    "Within shore position"
  )
)
```

```{r}
#| label: plot-jaccard-comparisons-regions-18S

compare_jaccard_regions_18S <- ggplot(filtered_sites_18S, aes(x = Var1_country, y = value, colour = comparison, fill = comparison)) +
  geom_boxplot(alpha = 0.5) +
  #geom_point(position = position_jitterdodge(dodge.width = .5), alpha = 0.2) +
  #geom_text(data = averages, aes(x = Var1_site, y = mean_value, label = round(mean_value, 2)),
  #vjust = -1.8, color = "black", size = 3.5, position = position_dodge(width = 0.75)) +
  labs(x = "",
       y = "Jaccard Dissimilarity (18S)") +
  scale_fill_manual(values = c("HighLow" = "#7DCE82", 
                               "Within shore position" = "grey50",
                               "LowOpenWater" = "#FF8811",
                               "HighOpenWater" = "#5998C5"), 
                    labels = c("HighLow" = "High vs Low", 
                               "HighOpenWater" = "High vs Open Water",
                               "LowOpenWater" = "Low vs Open Water",
                               "Within shore position" = "Same shore height"),
                    name = "Shore Height Comparison") +
  scale_colour_manual(values = c("HighLow" = "green4", 
                                 "Within shore position" = "grey30",
                                 "LowOpenWater" = "orange3",
                                 "HighOpenWater" = "skyblue4"),
                      labels = c("HighLow" = "High vs Low", 
                                 "HighOpenWater" = "High vs Open Water",
                                 "LowOpenWater" = "Low vs Open Water",
                                 "Within shore position" = "Same shore height"),
                      name = "Shore Height Comparison")+
  theme_classic() +
  scale_x_discrete(limits = country_order) +
  theme(axis.text.x = element_blank(),
        legend.position = "none")   +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1))

compare_jaccard_regions_18S
```

```{r}
#| label: get-legend-for-plot

compare_jaccard <- ggplot(filtered_sites, aes(x = Var1_site, y = value, fill = comparison, colour = comparison)) +
  geom_boxplot(alpha = 0.5) +
  #geom_text(data = averages, aes(x = Var1_site, y = mean_value, label = round(mean_value, 2)),
            #vjust = -1.8, color = "black", size = 3.5, position = position_dodge(width = 0.75)) +
  labs(x = "Sites",
       y = "") +
  scale_fill_manual(values = c("HighLow" = "#7DCE82", 
                               "Within shore position" = "grey50",
                               "LowOpenWater" = "#FF8811",
                               "HighOpenWater" = "#5998C5"), 
                    labels = c("HighLow" = "High vs Low", 
                               "HighOpenWater" = "High vs Open Water",
                               "LowOpenWater" = "Low vs Open Water",
                               "Within shore position" = "Same shore height"),
                    name = "Shore height comparison") +
  scale_colour_manual(values = c("HighLow" = "green4", 
                                 "Within shore position" = "grey30",
                                 "LowOpenWater" = "orange3",
                                 "HighOpenWater" = "skyblue4"),
                      labels = c("HighLow" = "High vs Low", 
                                 "HighOpenWater" = "High vs Open Water",
                                 "LowOpenWater" = "Low vs Open Water",
                                 "Within shore position" = "Same shore height"),
                      name = "Shore height comparison")+
  theme_classic() +
  scale_x_discrete(limits = site_order_reduced) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 1, byrow = TRUE)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1))

g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
}

legend <- g_legend(compare_jaccard)
```

```{r}
#| label: plot-jaccard-comparisons-region-all

region_jaccard <- cowplot::plot_grid(
  cowplot::plot_grid(compare_jaccard_regions_18S, compare_jaccard_regions, ncol = 1, labels = c("a", "b"), rel_heights = c(0.8, 1)), legend, ncol =1, rel_heights = c(1, 0.06))

ggsave(filename = "Figures/compare_jaccard_regionsonly_bothprimers.png", plot = region_jaccard ,
       device = "png", dpi = 600, units = "mm", height = 190, width = 200)

region_jaccard 
```

#### Composition plots (shore height x site)

Let's take a closer look at what might be driving differences between similar shore heights.

```{r}
#| label: shoreheight-across-sites

# 1. Create output directory
output_dir <- "Figures/CompAcrossSitesSamples"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# 2. Extract site names (verbatimLocality must be a sample variable)
site_names <- unique(sample_data(phylo_rocky_eDNA)$verbatimLocality)

# loop
for (site in site_names) {
  
  # Subset phyloseq object by site
  site_pseq <- subset_samples(phylo_rocky_eDNA, verbatimLocality == site)
  
  # Transform to relative abundance (compositional)
  pseq_relabund <- microbiome::transform(site_pseq, "compositional")
  
  # Aggregate rare phyla
  phylum_rareaggregate_pseq <- microbiome::aggregate_rare(
    pseq_relabund,
    level = "phylum",
    detection = 0.01,
    prevalence = 5 / 100
  )
  
  # Melt to data frame
  df <- psmelt(phylum_rareaggregate_pseq)
  
  # Plot manually
  p <- ggplot(df, aes(x = Sample, y = Abundance, fill = phylum)) +
    geom_bar(stat = "identity") +
    facet_wrap(~shorePosition, scales = "free_x") +
    xlab("Sample") +
    ylab("Relative abundance") +
    ggtitle(paste("Phylum relative abundance -", site)) +
    scale_fill_viridis_d(option = "turbo") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  # Save plot
  ggsave(
    filename = file.path(output_dir, paste0("phylum_bar_", gsub("[^A-Za-z0-9]", "_", site), ".png")),
    plot = p,
    width = 10,
    height = 6
  )
}
```

## Ecological concordance

We need to filter our data for those which we have ecological information on. In our case, we are looking at the MarClim species.

```{r}
#| label: define-marclim-genera

Marclim_genera_list <- c(
  "Actinia",
  "Alaria",
  "Anemonia",
  "Ascophyllum",
  "Asparagopsis",
  "Asterias",
  "Asterocarpa",
  "Aulactinia",
  "Austrominius",
  "Balanus",
  "Bifurcaria",
  "Botrylloides",
  "Calliblepharis",
  "Calliostoma",
  "Caulacanthus",
  "Chondracanthus",
  "Chondrus",
  "Chthamalus",
  "Clibanarius",
  "Codium",
  "Colpomenia",
  "Corella",
  "Crepidula",
  "Dasysiphonia",
  "Dendrodoa",
  "Diadumene",
  "Dictyopteris",
  "Didemnum",
  "Ericaria",
  "Fucus",
  "Grateloupia",
  "Halichondria",
  "Halidrys",
  "Haliotis",
  "Hemigraphsus",
  "Hemigraspus",
  "Himanthalia",
  "Hymeniacidon",
  "Laminaria",
  "Leptasterias",
  "Lichina",
  "Littorina",
  "Magallana",
  "Mastocarpus",
  "Melarhaphe",
  "Mytilus",
  "Nucella",
  "Ocenebra",
  "Ocinebrellus",
  "Onchidella",
  "Padina",
  "Palmaria",
  "Paracentrotus",
  "Patella",
  "Pelvetia",
  "Perforatus",
  "Perophora",
  "Phorcus",
  "Pikea",
  "Pollicipes",
  "Sabellaria",
  "Saccharina",
  "Saccorhiza",
  "Sargassum",
  "Semibalanus",
  "Steromphala",
  "Strongylocentrotus",
  "Testudinalia",
  "Treptacantha",
  "Undaria",
  "Urosalpinx",
  "Watersipora"
)
```

Let's look at the number of reads of each MarClim species across shore heights, sites and regions. These will be saved into the figure folder within the project.

```{r}
#| label: marclim-taxa-shore-height-figures

for (i in Marclim_genera_list) {
  tryCatch({
    # Subset the taxa for the current genus
    taxa <- subset_taxa(phylo_rocky_eDNA, genus == i)
    
    # Check if there are any taxa after subsetting
    if (ntaxa(taxa) == 0) {
      stop(paste("No taxa found for genus:", i))  # Manually trigger an error if no taxa are found
    }
    
    # Generate the bar plot for the current genus
    p <- plot_bar(taxa, "shorePosition", fill = "scientificname", title = paste(i, "across samples")) + 
      labs(y = "Read count", x = "Shore Height") +
      theme_classic()
    p
    
    # Save the plot to the folder
    plot_file_path <- paste0("Figures/Marclim_Shoreheights/", i, "_shoreposition.png")
    ggsave(filename = plot_file_path, plot = p, width = 8, height = 6, dpi = 600)  # Adjust size as needed
    
  }, error = function(e) {
    # Handle the error: print a message and skip to the next genus
    message(paste("Error with genus:", i, " - ", e$message))
  })
}

```

```{r}
#| label: marclim-taxa-regions-figures

for (i in Marclim_genera_list) {
  tryCatch({
    # Subset the taxa for the current genus
    taxa <- subset_taxa(phylo_rocky_eDNA, genus == i)
    
    # Check if there are any taxa after subsetting
    if (ntaxa(taxa) == 0) {
      stop(paste("No taxa found for genus:", i))  # Manually trigger an error if no taxa are found
    }
    
    # Generate the bar plot for the current genus
    p <- plot_bar(taxa, "country", fill = "scientificname", title = paste(i, "across regions")) + 
      labs(y = "Read count", x = "Shore Height") +
      theme_classic()
    p
    
    # Save the plot to the folder
    plot_file_path <- paste0("Figures/Marclim_Regions/", i, "_regions.png")
    ggsave(filename = plot_file_path, plot = p, width = 11, height = 9, dpi = 600)  # Adjust size as needed
    
  }, error = function(e) {
    # Handle the error: print a message and skip to the next genus
    message(paste("Error with genus:", i, " - ", e$message))
  })
}
```

```{r}
#| label: marclim-taxa-sites-figures

for (i in Marclim_genera_list) {
  tryCatch({
    # Subset the taxa for the current genus
    taxa <- subset_taxa(phylo_rocky_eDNA, genus == i)
    
    # Check if there are any taxa after subsetting
    if (ntaxa(taxa) == 0) {
      stop(paste("No taxa found for genus:", i))  # Manually trigger an error if no taxa are found
    }
    
    # Generate the bar plot for the current genus
    p <- plot_bar(taxa, "verbatimLocality ", fill = "scientificname", title = paste(i, "across sites")) + 
      labs(y = "Read count", x = "Shore Height") +
      theme_classic() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    p
    
    # Save the plot to the folder
    plot_file_path <- paste0("Figures/Marclim_Sites/", i, "_sites.png")
    ggsave(filename = plot_file_path, plot = p, width = 11, height = 9, dpi = 600)  # Adjust size as needed
    
  }, error = function(e) {
    # Handle the error: print a message and skip to the next genus
    message(paste("Error with genus:", i, " - ", e$message))
  })
}

```

Now let's make some overall composition plots. We need to do some data wrangling first.

```{r}
#| label: adding-biogeo-data

# filter just Marclim species
eDNA_marclim <- phylo_rocky_eDNA_long %>%
  filter(genus %in% Marclim_genera_list)

# correct capitalisation
eDNA_marclim$taxa <- sapply(strsplit(eDNA_marclim$taxa, " "),
                            #split taxa into two
                            function(x)
                              paste(toupper(substr(x[1], 1, 1)) %>%  #converts genus to upper
                                      paste0(tolower(
                                        substr(x[1], 2, nchar(x[1]))
                                      )), tolower(paste(x[-1], collapse = " "))))

# add biogeo data
eDNA <- left_join(eDNA_marclim, biogeo, by = "taxa")
eDNA <- eDNA %>%
  mutate(pa = ifelse(reads > 0, 1, 0))

eDNA_present <- eDNA %>% filter(pa > 0)
```

### Shore height concordance

```{r}
#| label: wrangling-for-shore-height-comp

# Gather high, low, subtidal columns into a 'zone' column
df_long <- eDNA_present %>%
  pivot_longer(cols = c(high, low, subtidal),
               names_to = "zone",
               values_to = "in_zone") %>%
  filter(in_zone == TRUE)  # Only keep rows where the taxa is found in that zone

# Filter main dataset now that we've got new overall category
eDNA_present <- eDNA_present %>%
  filter(!is.na(true_category))  # Filter rows where true_category is not NA

# Summarize the data to count the number of taxa by shorePosition and zone
df_summary <- eDNA_present %>% #change this to df_long if want to use other zones
  dplyr::group_by(shorePosition, true_category) %>%
  dplyr::summarise(n_taxa = n(), .groups = 'drop')

# Calculate the total number of taxa in each shorePosition
df_summary <- df_summary %>%
  dplyr::group_by(shorePosition) %>%
  dplyr::mutate(total_taxa = sum(n_taxa)) %>%
  ungroup()

# Calculate the proportion of taxa in each zone within each shorePosition
df_summary <- df_summary %>%
  mutate(proportion = n_taxa / total_taxa)
```

Now plot the figure.

```{r}
#| label: shore-height-concordance-plot

shore_height_prop <- ggplot(df_summary,
                            aes(x = shorePosition, y = proportion, fill = true_category)) +
  geom_bar(stat = "identity") +
  labs(x = "Shore Height", y = "Proportion of Present Taxa", fill = "Recorded Shore Height") +
  scale_fill_manual(
    values = c(
      "high" = highshore_colour,
      "low" = lowshore_colour,
      "subtidal" = "grey70",
      "high-low" = "orange3",
      "low-open" = openwater_colour
    ),
    labels = c(
      "high" = "High only",
      "low" = "Low only",
      "subtidal" = "Open only",
      "high-low" = "High & Low",
      "low-open" = "Low & Open"
    )
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", proportion * 100)), # Format with 1 decimal place and add "%"
    position = position_stack(vjust = 0.5), # Center text within bars
    size = 3.5,
    color = "black"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

shore_height_prop
```

### Regional concordance

```{r}
#| label: wrangling-for-regional-comp

df_summary_region <- df_long %>%
  dplyr::group_by(country, biogeography) %>%
  dplyr::summarise(n_taxa = n(), .groups = 'drop')

list.of.sw.invasives <- df_long %>%  filter(biogeography == 'invasive', 
                                            country == 'Southwest England',
                                            reads > 0)
list.of.sw.invasives  <- unique(list.of.sw.invasives$scientificname)

list.of.sch.boreal <- df_long %>%  filter(biogeography == 'boreal', 
                                            country == 'Scotland',
                                            reads > 0)
list.of.sch.boreal   <- unique(list.of.sch.boreal$scientificname)

# Calculate the total number of taxa in each region
df_summary_region <- df_summary_region %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(total_taxa = sum(n_taxa)) %>%
  ungroup()

# Calculate the proportion of taxa in each zone within each shorePosition
df_summary_region <- df_summary_region %>%
  mutate(proportion = n_taxa / total_taxa)
```

Now plot the figure.

```{r}
#| label: regional-concordance-plot

region_biogeo_prop <- ggplot(df_summary_region,
                             aes(x = country, y = proportion, fill = biogeography)) +
  geom_bar(stat = "identity") +
  labs(x = "Region", y = "Proportion of Present Taxa", fill = "Recorded Biogeography") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c(
      "boreal" = "#9DBBAE",
      "lusitanian" = "#E2DBBE",
      "invasive" = "red4",
      "cosmopolitan" = "#188FA7"
    ),
    labels = c(
      "boreal" = "Boreal",
      "lusitanian" = "Lusitanian",
      "invasive" = "Invasive",
      "cosmopolitan" = "Cosmopolitan"
    )
  ) +
  geom_text(
    aes(label = ifelse(
      proportion * 100 >= 1.1, sprintf("%.1f%%", proportion * 100), ""
    )),
    # Hide labels <1%
    position = position_stack(vjust = 0.5),
    # Center text within bars
    size = 4,
    color = "black"
  ) +
  theme_classic() +
  scale_x_discrete(limits = country_order) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

region_biogeo_prop
```

Make overall figure.

```{r}
#| label: join-concordance-plot

concordance_plot_prop <- cowplot::plot_grid(
  shore_height_prop,
  region_biogeo_prop,
  labels = c("a", "b"),
  ncol = 2,
  align = "hv"
)
concordance_plot_prop 

ggsave(
  filename = "Figures/concordance_plot_prop.png",
  plot = concordance_plot_prop,
  device = "png",
  dpi = 600,
  units = "mm",
  height = 180,
  width = 310
)
```

## Concluding remarks

We have now created all figures and reproduced all analyses presented in the manuscript.
